Investigation of thermal reactionomes reveals stress tolerance vs. resistance in

### 2014-08-26 doc/ApTranscriptome_manuscript.Rmd
more edits to Intro

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].

We compared the thermal reactionomes of these species to gain insight on the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. First, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the same genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this pattern.

Second, spatially-divergent adaptive responses to temperature may involve changes

the molecular toolkit involved in short-term acclimation to temperature is the same as

finding that more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, which might rely to a greater extent on proteome stability [@ghosh2010] and behavioral avoidance mechanisms [@sunday2014], while the opposite would be true for cold temperatures.

Second, responsive genes may change in their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero2011]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes. Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance. Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes

Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011]. To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

>>>>

Here, we present a novel method for RNAseq analysis: the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally-responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. Ants are ecologically-important species involved in multiple symbioses¸ and are key contributors to seed dispersal [@lengyel2009; @rodriguez-cabal2012], litter processing and decomposition [@lyford1963; @folgarait1998; @terborgh2001; @holec2006; @wardle2010].

We compared the thermal reactionomes of these species to gain insight on the relationship between short-term physiological acclimation and longer-term adaptive changes to temperature. First, through the process of genetic assimilation [@pigliucci2006], adaptive responses are hypothesized to involve the same molecular pathways as short-term acclimatory responses. Finding that the genes up-regulated in response to cold temperatures in *A. carolinensis* have increased basal expression in the colder-climate species *A. picea*, and vice versa, would be consistent with this pattern. Second, species from more stressful environments may rely more upon constitutive tolerance than plastically-induced resistance mechanisms [@barshis2013]. Finding that fewer genes are up-regulated to high temperatures in the warmer-climate species *A. carolinensis* would support this hypothesis. Third, spatially-divergent selection often results in local adaptation (cite...). For gene expression, this local adaptation may be manifested through changes in the sensitivity of responsive genes to changes in temperature. Evidence for local adaptation in gene expression would be observed in heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea*, and conversely, that cold-responsive genes are activated at colder temperatures in *A. picea*.

Importantly, none of the above predictions depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].

To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.
